Pyg4ometry: a Python library for the creation of Monte Carlo radiation transport physical geometries

### 2020-08-30 public/paper/pyg4ometry.tex
Paper update

\section{Computer descriptions of geometry} \label{sec:geometric}
Central to a computer readable geometry of a system is how a solid is defined in three dimensions. There are numerous different ways to describe a
geometry, including constructive solid geometry, boundary representation and tessellated polygons, these are described briefly in the following sections.  
The Geant4 geometry specification is a mixture of geometry modelling techniques and described in detail.

%%%%%%%%%%%%%%%%% Constructive solid geometry (CSG) 
Constructive solid geometry uses boolean operations, subtraction, intersection and union, between simple solids shapes (e.g. cube, cylinder, sphere etc) to model complex surfaces which represent a solid. Boolean operations and solids can be combined to form a CSG-tree to model complex geometry. FLUKA uses CSG to 
model solids, the form of boolean expression used by FLUKA is not a general CSG-tree but a logical expression in disjunctive normal-form.

%%%%%%%%%%%%%%%%% Boundary representation (BREP)
Boundary representation (BREP) consists of two parts, topology and geometry. Topological elements are faces, edges and vertices. No current RTMC 
applications use native file formats employed by CAD systems. The conversion of CAD BREP formats for loading in RTMC applications is typically performed 
via a tessellated format, although it is possible to decompose BREP descriptions to simple half-spaces and solids as used in CSG descriptions.

@@

\subsection{Input and output} 
For each input format available in pyg4ometry a dedicated {\it Reader} class is implemented, so \verb|gdml.Reader|, \verb|stl.Reader|, \verb|fluka.Reader| 
and \verb|freecad.Reader|. Each reader constructs the appropriate Geant4 classes and provides a registry which can be used or manipulated by the user. 
Output consists of taking the registry and writing the appropriate GDML.

\subsection{Expression evaluation}
In GDML symbolic expressions can be used to parametrise solids and their placement. These expressions are evaluated when the GDML is loaded into Geant4. 
In order to fully replicate the functionality of GDML an expression engine was implemented using ANTLR \cite{10.5555/2501720}. GDML allows for the definition 
and assignment of variables. GDML expressions are not much more complicated than binary operators $+, -, \times, /$ and common trigonometric and special functions $\sin, \cos, \tan$ etc. The AST terminates on either expressions which evaluate to numbers or GDML variables. Internally all classes use GDML 
expressions and not floating point numbers. Storing internal data as expressions allows for deferred evaluation (or re-evaluation) of solid parameters and 
placements. This allows a user to update variables whilst defining geometry and the expression engine will update all internal values.

\subsection{Replica, division, parametrised and looped volumes}
A powerful feature of Geant4 and hence GDML is the ability to either repeat, divide or parametrise geometry. The create multiple replicas of a volume in a cartesian, cylindrical or spherical grid is known as a ReplicaVolume. Currently GDML loops are not supported by pyg4ometry and will be implemented in a future release.

\subsection{Internal data representation} 
The internal data representation follows closely the structure of GDML. A \verb|Registry| class aggregates python ordered dictionaries that are  used to store the main 
elements of a GDML file. As a user specifies geometry the registry is updated. When a user is finished with the geometry, the registry is written to disk as a GDML file.
\verb|Registry| instances can be aggregated to form a composite geometry or volumes can be removed and added.

@@

\subsection{Volume bounding} 
If a user is creating a placing multiple {\em daughter} volumes within a {\em mother} volume then it is the users responsibility to create a solid which fully 
encompasses the daughter volumes. Overlaps between daughter volumes and the mother can be detected, but it is desirable to have a mother volume 
shape that efficiently holds all of its daughters.

@@

\subsection{Adding properties to volumes}
A pure geometry description is an insufficient description for a RT simulation. Often other attributes need to be assigned for example material definition, magnetic field or  
optical surface properties.

\subsection{Overlap detection}
All RT transport codes can not handle spatial overlap between two geometric objects and likely are to have ill defined behaviour when tracking particles  
through such a situation.  A key feature of pyg4ometry is the detection of potential overlaps efficiently, it does this by performing a CSG intersection between solid instances 
and determining if the resulting mesh is empty. Generally the overlap detection algorithm proceeds as

{\small 
\begin{verbatim}
Overlap detection
\end{verbatim}
}

For a logical volume with $n_{\rm daughters}$ physical volumes, assuming meshes for solids with average number of vertices $n$, clearly this algorithm 
has complexity $\sim \mathcal{O}(n^2 n_{\rm daughters})$. This might seem like a prohibitive computational cost, but worth it considering the potential 
waste if small overlaps are present in the final RT simulation. This algorithm clearly favours geometry descriptions which have a high degree of logical 
volume reuse. Due to the discrete nature of triangular meshes it is not possible to have perfect detection of overlaps, especially when curved surfaces are 
considered. 
 \begin{figure}[htbp]
\begin{center}
\includegraphics[width=8cm]{./diagrams/overlap.pdf}
\caption{Placement of physical volume inside a logical volume}
\label{fig:overlap}
\end{center}
\end{figure}

\subsection{Graphical user interface}
A basic Qt based graphical user interface is also available with pyg4ometry. If a model has a large number of volumes defined in a hierarchical tree structure, it is cumbersome for the user to find the logical or physical volume.

@@

\subsection{Infinite bodies}

The majority of bodies in FLUKA are infinite in extent, and fall
broadly into four categories, half-spaces, infinitely long cylinders,
infinitely long elliptical cylinders and quadric surfaces.
Translating these bodies to Geant4 requires generating the equivalent
finite solid whilst retaining the same final finite Boolean shape.
This is achieved with the use of axis-aligned bounding boxes (AABB),
in which the FLUKA body is translated to a finite solid with with
dimensions slightly larger than the AABB.  The lengths of infinite
(elliptical) cylinders are reduced to finite equivalents with lengths
slightly greater than the bounding box.  Similarly, half spaces are
reduced to boxes with one face acting as that of the half-space face,
and quadric surfaces are sampled only over the volume denoted by the
AABB.  Furthermore, the positions of these solids are moved as close
to the bounding box as possible whilst retaining an identical final
Boolean geometry.

@@

\subsection{Removing redundant half-spaces}

The above algorithm for replacing infinite FLUKA bodies with finite
Geant4 solids works well in most cases, but additional care must be
taken for redundant infinite half spaces.  A redundant infinite half
space is defined as one which has no affect on the topology of the
final Boolean solid.  Whilst this may be true of arbitrary bodies, it
is most problematic for half spaces as after the infinite-body
reduction has been performed, if the half-space is far away from the
region's AABB, then it can result in a malformed Boolean.  This is
demonstrated in Figure~\ref{fig:redundant-halfspace}.  Such
half-spaces are filtered from their respective regions during the
conversion process by calculating the nearest distance from the centre
of the AABB to the half-space face.  If this distance is greater than
the centre-to-corner distance of the AABB, then that half-space is
removed from the region during conversion.

\subsection{Coplanar faces}

Coplanar faces, in which the faces of two union components or that of
two regions are perfectly coplanar in FLUKA are ubiquitous and present
no difficulties to the operation of the program.  However, in Geant4
these will generally result in tracking errors.  These must be handled
robustly to ensure the resulting geometry is usable.  Coplanar faces
are resolved automatically by slightly decreasing the size of every
body that is used in an intersection, and increasing the size of every
body used in a subtraction.  These rules are inverted for nested
subtractions and work well for guaranteeing well-formed geometry that
is free from overlaps.

>>>>

\section{Computer descriptions of geometry} \label{sec:geometric}
Central to a computer readable geometry of a system is how a solid is defined in three dimensions. There are numerous different ways to describe a
geometry, including constructive solid geometry, boundary representation and tessellated polygons, these are described briefly in the following sections.  
The Geant4 geometry specification is a mixture of geometry modelling techniques and described in detail last.

%%%%%%%%%%%%%%%%% Constructive solid geometry (CSG) 
Constructive solid geometry uses boolean operations, subtraction, intersection and union, between simple solid shapes (e.g. cube, cylinder, sphere etc) or infinite
volumes (e.g a plane defined infinite half-space) to model complex surfaces which represent a solid. Boolean operations and solids can be combined to form a 
CSG-tree to model complex geometry. FLUKA uses CSG to model solids, the form of boolean expression used by FLUKA is not a general CSG-tree but a 
logical expression in disjunctive normal-form.

%%%%%%%%%%%%%%%%% Boundary representation (BREP)
Boundary representation (BREP) consists of two parts, topology and geometry. Topological elements are faces, edges and vertices and the corresponding
geometrical elements are surfaces, curves and points. No current RTMC applications use native file formats employed by CAD systems. The conversion of CAD BREP 
formats for loading in RTMC applications is typically performed via a tessellated format, although it is possible to decompose BREP descriptions to bounded or 
infinite mathematical surfaces and subsequently solids as used in CSG descriptions.

@@

%%%%%%%%%%%%%%%%%  Input and output  
For each input format available in pyg4ometry a dedicated {\it Reader} class is implemented, so \verb|gdml.Reader|, \verb|stl.Reader|, \verb|fluka.Reader| 
and \verb|freecad.Reader|. Each reader constructs the appropriate Geant4 classes and provides a registry which can be used or manipulated by the user. 
Output consists of taking the registry and writing the appropriate GDML.

%%%%%%%%%%%%%%%%%  Expression evaluation
In GDML symbolic expressions can be used to parametrise solids and their placement. These expressions are evaluated when the GDML is loaded into Geant4. 
In order to fully replicate the functionality of GDML an expression engine was implemented using ANTLR \cite{10.5555/2501720}. GDML allows for the definition 
and assignment of variables. GDML expressions are not much more complicated than binary operators $+, -, \times, /$ and common trigonometric and special functions $\sin, \cos, \tan$ etc. The AST terminates on either expressions which evaluate to numbers or GDML variables. Internally all classes use GDML 
expressions and not floating point numbers. Storing internal data as expressions allows for deferred evaluation (or re-evaluation) of solid parameters and 
placements. This allows a user to update variables whilst defining geometry and the expression engine will update all internal values.

%%%%%%%%%%%%%%%%%  Replica, division, parametrised and looped volumes
A powerful feature of Geant4 and hence GDML is the ability to either repeat, divide or parametrise geometry. The create multiple replicas of a volume in a cartesian, cylindrical or spherical grid is known as a ReplicaVolume. A DivisionVolume breaks a primitive into segments in either a cartesian or cylindrical polar coordinates. 
A Parametrised Volume allows for the arbitrary multiple placement of solids where the parameters are allowed to vary for each placement.  GDML supports loops, 
where sections of GDML are allowed to be repeated with varying parameters. GDML loops are not supported by pyg4ometry and will be implemented in a 
future release.

%%%%%%%%%%%%%%%%%  Internal data representation
The internal data representation follows closely the structure of GDML. A \verb|Registry| class aggregates python ordered dictionaries that are  used to store the main 
elements of a GDML file. As a user specifies geometry the registry is updated. When a user is finished with the geometry, the registry is written to disk as a GDML file.
\verb|Registry| instances can be aggregated to form a composite geometry or volumes can be removed and added.

@@

%%%%%%%%%%%%%%%%% Volume bounding
If a user is creating a placing multiple {\em daughter} volumes within a {\em mother} volume then it is the users responsibility to create a solid which fully 
encompasses the daughter volumes. Overlaps between daughter volumes and the mother can be detected, but it is desirable to have a mother volume 
shape that efficiently holds all of its daughters.

@@

%%%%%%%%%%%%%%%%% Adding properties to volumes
A pure geometry description is an insufficient description for a RT simulation. Often other attributes need to be assigned for example material definition, magnetic field or  
optical surface properties.

\subsection{Overlap detection}
All RT transport codes can not handle spatial overlap between two geometric objects and likely are to have ill defined behaviour when tracking particles  
through such a situation.  A key feature of pyg4ometry is the detection of potential overlaps efficiently, it does this by performing an intersection operation etween solid instances and determining if the resulting mesh is empty. Figure~\ref{fig:overlap} shows the potential overlaps, a) protrusion of a daughter from the mother, b) and intersection with finite volume and c) an overlap where two daughters share a face. 
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=8cm]{./diagrams/overlap.pdf}
\caption{Schematic of the three different types of overlap between daughters of a mother logical volume.}
\label{fig:overlap}
\end{center}
\end{figure}

Overlap detection in pyg4ometry relies of the meshes generated for each solid, any detection algorithm will be approximate. Generally the overlap detection 
algorithm proceeds as 
\begin{enumerate}
\item Check if any of the daughter solid meshes have a non-null intersection with the compliment of mother solid mesh
\item Loop over all daughter volumes and check
\item Check
\end{enumerate}

For a logical volume with $n_{\rm daughters}$ physical volumes, assuming meshes for solids with average number of vertices $n$, clearly this algorithm 
has complexity $\sim \mathcal{O}(n^2 n_{\rm daughters})$. This might seem like a prohibitive computational cost, but worth it considering the potential 
waste if small overlaps are present in the final RTMC simulation. This algorithm clearly favours geometry descriptions which have a high degree of logical 
volume reuse. Due to the discrete nature of triangular meshes it is not possible to have perfect detection of overlaps, especially when curved surfaces are 
considered.

%%%%%%%%%%%%%%%%% Graphical user interface
% A basic Qt based graphical user interface is also available with pyg4ometry. If a model has a large number of volumes defined in a hierarchical tree structure, it is % cumbersome for the user to find the logical or physical volume.

@@

\subsection{Infinite bodies}
The majority of bodies in FLUKA are infinite in extent, and fall
broadly into four categories, half-spaces, infinitely long cylinders,
infinitely long elliptical cylinders and quadric surfaces.
Translating these bodies to Geant4 requires generating the equivalent
finite solid whilst retaining the same final finite Boolean shape.
This is achieved with the use of axis-aligned bounding boxes (AABB),
in which the FLUKA body is translated to a finite solid with with
dimensions slightly larger than the AABB.  The lengths of infinite
(elliptical) cylinders are reduced to finite equivalents with lengths
slightly greater than the bounding box.  Similarly, half spaces are
reduced to boxes with one face acting as that of the half-space face,
and quadric surfaces are sampled only over the volume denoted by the
AABB.  Furthermore, the positions of these solids are moved as close
to the bounding box as possible whilst retaining an identical final
Boolean geometry.

@@

\subsection{Removing redundant half-spaces}
The above algorithm for replacing infinite FLUKA bodies with finite
Geant4 solids works well in most cases, but additional care must be
taken for redundant infinite half spaces.  A redundant infinite half
space is defined as one which has no affect on the topology of the
final Boolean solid.  Whilst this may be true of arbitrary bodies, it
is most problematic for half spaces as after the infinite-body
reduction has been performed, if the half-space is far away from the
region's AABB, then it can result in a malformed Boolean.  This is
demonstrated in Figure~\ref{fig:redundant-halfspace}.  Such
half-spaces are filtered from their respective regions during the
conversion process by calculating the nearest distance from the centre
of the AABB to the half-space face.  If this distance is greater than
the centre-to-corner distance of the AABB, then that half-space is
removed from the region during conversion.

\subsection{Coplanar faces}
Coplanar faces, in which the faces of two union components or that of
two regions are perfectly coplanar in FLUKA are ubiquitous and present
no difficulties to the operation of the program.  However, in Geant4
these will generally result in tracking errors.  These must be handled
robustly to ensure the resulting geometry is usable.  Coplanar faces
are resolved automatically by slightly decreasing the size of every
body that is used in an intersection, and increasing the size of every
body used in a subtraction.  These rules are inverted for nested
subtractions and work well for guaranteeing well-formed geometry that
is free from overlaps.
